Parametrization of Invertible Resnets up to Homology Equivalence for Abelian Lie Groups

### 2020-04-02 llncs2e/paper77.tex
clarification

\section{Motivation}
Neural nets and their underlying principles were studied from the perspective of differential and information geometry \citemain{YangA97}. The natural gradient descent turned out to be Fisher optimal \citemain{Amari1,Amari2}, in its interpretation as a walk on the hypersurface of a neural loss function. Local minima, maxima and saddle points on the hypersurface have always been a problem when training machine learning models. A critical point on this surface corresponding to the global minimum of a smaller model forms a local minimum or saddle point of a larger one \citemain{FukumizuA00}. In other words, the critical values of a function represented by a layer of $(n-1)$-perceptrons will result in many critical values of a layer of $n$-perceptrons. This motivates an appropriately sized model.

@@

The idea originates from Lie theory. We use invertible architectures, to guarantee our assumptions and apply ascending sorting on the samples to yield permutation invariance or abelianization of the input domain \citemain{ArdizzoneKRK19,NguyenAK19}. Our interpretation of a neural net is a continuous transformation, transferring invariant structures of the data in the input space to its deepest layer. Thus, it is capable to reconstruct the structure of the original space. This approach corresponds to the concept of homotopy equivalence \citemain{RubinsteinR08,YangY93}, which means that there is a continuous map $f: X \rightarrow Y$ and $g: Y \rightarrow X$, such that $f \circ g$ is homotopic to the identity $\text{Id}_Y$ and $g \circ f$ to $\text{Id}_X$. Thus, the spaces can be deformed continuously into one another. To do so, their homotopy groups must be equal. The zeroth homotopy group gives information about connected components which can be collapsed to a point by continuous functions, the first homotopy group about `holes' and the second about `cavities' and so on. But homotopies are difficult to compute \citemain{klaus2004quick,hatcher2002algebraic}. Homology theory gives us a remedy that guarantees an isomorphism between homotopy groups and homology groups under the conditions of Hurewicz's Theorem \citemain{klaus2004quick}. We use theoretical results from Lie theory and persistent homology theory to predict how many dimensions of embedding are needed to represent the homology groups. Our paper is structured as follows:
\begin{itemize}
    \item In Sect. \ref{sec:manifoldassumption} we discuss the mathematical assumptions and recall the principles of smooth manifolds and Lie theory. In particular, we recall the decomposition of connected commutative Lie groups into $\mathbb{R}^p \times \mathbb{T}^q$.
    \item In Sect. \ref{sec:persistenthomology} we introduce to the theory of persistent homology, define persistence diagrams, persistence landscapes and confidence sets, in order to estimate homology groups despite noisy data. We use Künneth's theorem and derive how to count the homology groups of the input space.
    \item In Sect. \ref{sec:neuralnettheory} we explain the mathematical theory of neural nets and construct invertible deep neural nets to satisfy our assumptions. Resnets, which have a larger state space dimension, offer an architecture with a comparatively small number of parameters to dense nets.
    \item In Sect. \ref{sec:numericalexperiments} we validate our theory training autoencoders on the cifar10 and cifar100 datasets, isolating the innermost layer such that no residual connections bypass it on different embedding dimensions.
\end{itemize}

@@

\paragraph{Neural Nets} The data manifold is supposed to be a commutative Lie group. Thus, we artificially generate commutativity by sorting input and target vectors beforehand. We perturb each input vector $\textbf{x}^{(0)}$ with Gaussian noise, such that the input is weighted $0.8 \cdot \textbf{x}^{(0)} + 0.5 \cdot \epsilon$ with $\epsilon \sim \mathcal{N}(0,\sigma^2 \mathbf{I})$. These input vectors are trained to the sorted vector $\textbf{x}^{(0)}$. We use $\text{L}(\textbf{x}^{(0)}, \textbf{y}) = \frac{1}{n} \sum_{i=1}^{n} \left(\textbf{x}^{(0)}_i-\textbf{y}_i\right)^2$ as loss function. The sigmoid activation function $\sigma(\mathbf{x}^{(l)}) = \frac{1}{1 + e^{-\mathbf{x}^{(l)}}}$ is applied as elementwise activation function throughout the net. In all experiments dense neural nets are used with a bias term. The datasets are flattened from $(32,32,3)$ into $(3072)$. The first layer is applied to the flattened dataset, which is also adapted by back propagation, i.e. $\varphi^{(0)}\left(x^{(0)} \right) =$ $\sigma\left( \sum_{i=1}^{n} \textbf{W}_{ji}^{(1)} \textbf{x}_{i}^{(0)} + \textbf{b}_{i}^{(1)} \right)$. It transfers the data to the desired embedding dimension. The subsequent layers are implemented as $\mathcal{C}^1$-residual invertible layers, see Sect. \ref{resnet} and Sect. \ref{subsec:invertibleneuralnets}. A total of $5$ of these hidden layers are used, all with the same embedding dimension. More details can be found in Appx. \ref{appx:numexp}.

\subsection{Results}
Tab. \ref{stats} shows the representatives of homology classes from their respective groups for cifar10 and cifar100. Further, it shows the estimated minimal embedding dimension for a manifold with these homology groups for $\mathcal{C}^1$ to $\mathcal{C}^4$-resnets. The representatives for the $k$-th homology group $\sharp_k^{H}\left(\bar{\mathfrak{L}}\left(\mathfrak{P}(K(X))\right)\right)$ of the respective dataset are counted. Then, we seek for an embedding dimension to represent a manifold with the number of spheres $q$ with these homology groups. A Monte Carlo approach is used to find integer solutions of Eq. \ref{countingequation}. All models below our dimensional treshold show explosions of the gradient, see Fig. \ref{costfunctions}. The dimension of the respective loss function can be identified by the transition \gradient{}, from the lowest tested dimension to the highest, see Fig. \ref{costfunctions}. The models above the treshold remain stable. Higher parameterized tested models are stable. A single net still shows an explosive gradient in Fig. \ref{costfunctions} (b). This could be due to statistical errors in parametrization by mean confidence persistence landscapes. The net showing the deflection of gradient is close to our threshold. We specify the parameterization for an embedding as well as for neural nets of the smoothness degree $1$--$4$. The noise was added after sorting. The neural net is capable of eliminating the additional structure of noise, which is no longer part of the abelian Lie group.
\input{figure4.tex}

\section{Conclusions}
\label{sec:conclusion}
Based on the theory of Lie groups and persistent homology a method for invertible resnets was developed to parameterize them with good precision due to abelianization of the input domain. This forms a solution for a special case of learning problems, allowing to restrict the domain to a commutative group. The initial problem could be simplified by decomposing the Lie group into a product space $\mathbb{R}^p \times \mathbb{T}^q$. The homology groups of the $q$-torus and the Euclidean space are well known. The application of fully invertible neural nets to resnets demonstrates our results on a different Lie group from Euclidean space. By applying Künneth's theorem the homology groups of the topological factor spaces could be connected with the ones of their product space. Using mean confident persistence landscapes the homology groups were estimated. A treshold was computed, see sect. \ref{sec:numericalexperiments}, which is determined by the state space dimension of the resets according to a representation up to homology equivalence. The mean persistence landscapes are particularly useful for short-lived isolated collections of data points, which in particular generate elements from homology classes at the beginning of the filtering and become visible at the lower left corner of the persistence landscapes, see Fig. \ref{homgroups} (a). The theoretical work was tested with numerical experiments on toy data. We do not limit the processing of the neural net within the layers to an abelian group. Instead, we assume that the net learns this structure from the input domain. This seems to hold, according to our experiments. Still, it can cause inaccuracies in the estimation of the dimensional treshold. The abelianization of the invertible layers would solve this problem.

>>>>

\section{Motivation}
Neural nets and their underlying principles were studied from the perspective of differential and information geometry \citemain{YangA97}. The natural gradient descent turned out to be Fisher optimal \citemain{Amari1,Amari2}, in its interpretation as a walk on the hypersurface of a neural loss function. Local minima, maxima and saddle points on the hypersurface have always been a problem when training machine learning models. A critical point on this surface corresponding to the global minimum of a smaller model forms a local minimum or saddle point of a larger one \citemain{FukumizuA00}. In other words, the critical values of a function represented by a layer of $(n-1)$-perceptrons will result in many critical values of a layer of $n$-perceptrons.

@@

We use invertible architectures, to demonstrate our hypothesis on a different lie Group from Euclidean space \citemain{NguyenAK19}. Our interpretation of a neural net is a continuous transformation, transferring invariant structures of the data in the input space to its deepest layer. Thus, it is capable to reconstruct the structure of the original space. This approach corresponds to the concept of homotopy equivalence \citemain{RubinsteinR08,YangY93}, which means that there is a continuous map $f: X \rightarrow Y$ and $g: Y \rightarrow X$, such that $f \circ g$ is homotopic to the identity $\text{Id}_Y$ and $g \circ f$ to $\text{Id}_X$. Thus, the spaces can be deformed continuously into one another. To do so, their homotopy groups must be equal. The fundamental group gives information about connected components which can be collapsed to a point by continuous functions, the first homotopy group about `holes' and the second about `cavities' and so forth. But homotopies are difficult to compute \citemain{klaus2004quick,hatcher2002algebraic}. Homology theory gives a remedy that guarantees an isomorphism between homotopy groups and homology groups under the conditions of Hurewicz's Theorem \citemain{klaus2004quick}. We use theoretical results from Lie theory and persistent homology theory to estimate how many dimensions of embedding are needed to represent the homology groups. Our paper is structured as follows:
\begin{itemize}
    \item In Sect. \ref{sec:manifoldassumption} we discuss the mathematical assumptions and recall the principles of smooth manifolds and Lie theory. In particular, we recall the decomposition of connected commutative Lie groups into $\mathbb{R}^p \times \mathbb{T}^q$.
    \item In Sect. \ref{sec:persistenthomology} we introduce to the theory of persistent homology, define persistence diagrams, persistence landscapes and confidence sets, in order to estimate homology groups despite noisy data. We use Künneth's theorem and derive how to count the homology groups of the input space.
    \item In Sect. \ref{sec:neuralnettheory} we explain the mathematical theory of neural nets and construct invertible deep neural nets to satisfy our assumptions. Resnets, which have a larger state space dimension, offer an architecture with a comparatively small number of parameters to dense nets.
    \item In Sect. \ref{sec:numericalexperiments} we validate our theory training autoencoders on the cifar10 and cifar100 datasets, isolating the innermost layer such that no residual connections bypass it on different embedding dimensions.
\end{itemize}

@@

\paragraph{Neural Nets} The data manifold is supposed to be some commutative Lie group. We perturb each input vector $\textbf{x}^{(0)}$ with Gaussian noise, such that the input is weighted $0.8 \cdot \textbf{x}^{(0)} + 0.5 \cdot \epsilon$ with $\epsilon \sim \mathcal{N}(0,\sigma^2 \mathbf{I})$. These input vectors are trained to the original vector $\textbf{x}^{(0)}$. We use $\text{L}(\textbf{x}^{(0)}, \textbf{y}) = \frac{1}{n} \sum_{i=1}^{n} \left(\textbf{x}^{(0)}_i-\textbf{y}_i\right)^2$ as loss function. The sigmoid activation function $\sigma(\mathbf{x}^{(l)}) = \frac{1}{1 + e^{-\mathbf{x}^{(l)}}}$ is applied as elementwise activation function throughout the net. In all experiments dense neural nets are used with a bias term. The datasets are flattened from $(32,32,3)$ into $(3072)$. The first layer is applied to the flattened dataset, which is also adapted by back propagation, i.e. $\varphi^{(0)}\left(x^{(0)} \right) =$ $\sigma\left( \sum_{i=1}^{n} \textbf{W}_{ji}^{(1)} \textbf{x}_{i}^{(0)} + \textbf{b}_{i}^{(1)} \right)$. It transfers the data to the desired embedding dimension. The subsequent layers are implemented as $\mathcal{C}^1$-residual invertible layers, see Sect. \ref{resnet} and Sect. \ref{subsec:invertibleneuralnets}. A total of $5$ of these hidden layers are used, all with the same embedding dimension. More details can be found in Appx. \ref{appx:numexp}.

\subsection{Results}
Tab. \ref{stats} shows the representatives of homology classes from their respective groups for cifar10 and cifar100. Further, it shows the estimated minimal embedding dimension for a manifold with these homology groups for $\mathcal{C}^1$ to $\mathcal{C}^4$-resnets. The representatives for the $k$-th homology group $\sharp_k^{H}\left(\bar{\mathfrak{L}}\left(\mathfrak{P}(K(X))\right)\right)$ of the respective dataset are counted. Then, we seek for an embedding dimension to represent a manifold with the number of spheres $q$ with these homology groups. A Monte Carlo approach is used to find integer solutions of Eq. \ref{countingequation}. All models below our dimensional treshold show explosions of the gradient, see Fig. \ref{costfunctions}. The dimension of the respective loss function can be identified by the transition \gradient{}, from the lowest tested dimension to the highest, see Fig. \ref{costfunctions}. The models above the treshold remain stable. Higher parameterized tested models are stable. A single net still shows an explosive gradient in Fig. \ref{costfunctions} (b). This could be due to statistical errors in parametrization by mean confidence persistence landscapes. The net showing the deflection of gradient is close to our threshold. We specify the parameterization for an embedding as well as for neural nets of the smoothness degree $1$--$4$. The neural net is capable of eliminating the additional structure of noise, which is no longer part of the abelian Lie group.
\input{figure4.tex}

\section{Conclusions}
\label{sec:conclusion}
Based on the theory of Lie groups and persistent homology a method for invertible resnets was developed to parameterize them with good precision due to the assumption that the data lies on or near by some commutative Lie group. This forms an approximate solution for a special case of learning problems, allowing to restrict the domain to a commutative group. The initial problem could be simplified by decomposing the Lie group into a product space $\mathbb{R}^p \times \mathbb{T}^q$. The homology groups of the $q$-torus and the Euclidean space are well known. The application of fully invertible neural nets to resnets demonstrates our results on a different Lie group from Euclidean space. By applying Künneth's theorem the homology groups of the topological factor spaces could be connected with the ones of their product space. Using mean confident persistence landscapes the homology groups were estimated. A treshold was computed, see sect. \ref{sec:numericalexperiments}, which is determined by the state space dimension of the resets according to a representation up to homology equivalence. The mean persistence landscapes are particularly useful for short-lived isolated collections of data points, which in particular generate elements from homology classes at the beginning of the filtering and become visible at the lower left corner of the persistence landscapes, see Fig. \ref{homgroups} (a). The theoretical work was tested with numerical experiments on toy data. We do not limit the processing of the neural net within the layers to an abelian group. Instead, we assume that the net learns this structure from the input domain. This seems to hold, according to our experiments. Still, it can cause inaccuracies in the estimation of the dimensional treshold. The abelianization of the invertible layers would solve this problem.
